The Effect of Weather Shocks on Children's Education in Bolivia

### 2020-05-09 final_seminar_paper.Rmd
adding more text from word

---
title: "The Effect of Weather Shocks on Children's Education in Bolivia"
author: "Katya Garcia-Israel"
abstract: "Weather shocks have the potential to dramatically affect households' livelihoods and income levels, especially in developing countries. A growing subset of developmental and environmental literature focuses on investigating the relationship between environment and weather factors and household changes in behavior. One frequently cited response to negative weather shocks that decrease income is an increase in child labor to supplement the household income and a decrease in school attendance. This mechanism is especially relevant in agricultural areas, but can also be seen in more urban areas if the weather events cause a loss of property, injury, or displace people. This study explores the relationship between flood and drought occurrences and school attendance in Bolivia. I find that ..."
header-includes:
   - \usepackage{booktabs}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage[labelfont=bf]{caption}
output: 
  pdf_document:
    number_sections: true
    keep_tex: true

@@

@@

@@

@@

>>>>

---
title: "The Effect of Weather Shocks on Children's Education in Bolivia"
author: "Katya Garcia-Israel"
abstract: "Weather shocks have the potential to dramatically affect households' livelihoods and income levels, especially in developing countries. A growing subset of developmental and environmental literature focuses on investigating the relationship between environment and weather factors and household changes in behavior. One frequently cited response to negative weather shocks that decrease income is an increase in child labor to supplement the household income and a decrease in school attendance. This mechanism is especially relevant in agricultural areas but can also be seen in more urban areas if the weather events cause a loss of property, injury, or displace people. This study explores the relationship between flood and drought occurrences and school attendance in Bolivia. Contrary to my hypothesis, I find that floods and droughts actually have positive impacts on school outcomes, a relationship that might indicate that weather events negatively impact agricultural harvests and decrease the demand for children’s labor, thereby allowing them to go to school."
header-includes:
   - \usepackage{booktabs}
   - \usepackage{setspace}
   - \doublespacing
   - \usepackage[labelfont=bf]{caption}
output: 
  pdf_document:
    number_sections: true
    keep_tex: true

@@

Weather events have been linked to changes in household income, education and working behaviors, especially in developing countries and areas where people are particularly reliant on agriculture. In the 21st century, with the threat of climate change, exploration of these connections is even more crucial. In developmental economics literature, the mechanisms by which weather events affect households’ livelihoods tend to be agricultural, linking adverse precipitation or temperature shocks to decreases in income. This paper seeks to answer the question of whether high occurrences of floods and droughts in a region have an effect on children’s education, specifically in Bolivia.

A key reason for choosing Bolivia as a case study for this research question is its extensive climactic diversity. Despite this high level of internal precipitation and temperature variation, the ecological zones correspond fairly well to political boundaries of the country, which is divided into nine departments. Three departments fall into each ecological category, which are mountains (La Paz, Potosi, and Oruro), valleys (Cochabamba, Tarija, Chuquisaca), and plains (Santa Cruz, Beni, Pando). Figure 1 shows the three regional climate divisions and the departmental boundaries, with the mountains in the west, the plains in the north and east, and the valleys in the south-central part of the country.

@@

Table 2 provides summary statistics for the variables used in the analysis using the full sample of children, and Table 3 shows the summary statistics for children living in rural areas. The rural variable in this analysis is based on whether the households live in a municipality with fewer than 2000 people. The total sample contains 24,316 observations while the rural sample is considerably smaller, with 9,804 observations. Many of the statistics have similar values in both samples, but it is notable more children are registered for school in the total sample than in the rural sample, as evidenced by the higher mean on the register variable. The average household income is also lower for children in the rural sample, while the percentage of households below the poverty line is higher. Overall, most of the children ages 5-17 do attend school, with the high mean of the attend variable (0.986).

@@

The following graphs plot the flood variables and the drought variables over the sample years to capture the variation between departments and within the same department over time. Figure 2 shows the flood occurrences over time, and Figure 3 shows the droughts. Many of the departments do not actually have much variation between the years, but some events stand out in the graph and are clearly extreme shocks. There is no evidence of a positive or a negative time trend in these variables. Comparing the two graphs, it is evident that in many of the departments have high counts of floods in the same year as low counts of droughts, instead of experiencing both shocks at once in different parts of the region. This is reassuring in terms of being able to assume that the departments are a decent measure of location, as some are quite large, but evidently not so internally ecologically diverse that it confuses the analysis. The main challenge of this data is the unit of location by which the observations are grouped, as the department level is the most disaggregated measure. Yet households on one side of a department many not be experiencing the full effects of the weather shocks than households in another location within the same department.

@@

The flood results ended up being contrary to the hypothesis discussed in the conceptual framework section, the theory supported by the literature review that weather events will have detrimental impacts on children’s education via the mechanism of agricultural damages requiring increased child labor. As the results were highly significant, it suggests that in the Bolivian context during the years from 2005-2012, there was another relationship between floods and droughts and schooling. The alternate theory is that weather events have negative effects on agricultural activities in an area, which then decreases the demand for labor in agriculture sectors.  If children’s labor is no longer required if crops are destroyed, the opportunity cost of going to school lessens considerably and they may become more likely to register and attend. The significance and direction of the coefficient on the interaction terms linking floods with droughts in the main probit regression also lend credence to that theory.

The drought results for the rural sample did correspond to the argument that weather events decrease income and increase the opportunity cost of school attendance as there was a significant negative relationship between droughts and school attendance. The result in Table 4 assumes that the effect of floods or droughts on schooling will be seen in the same year. This is a more convincing assumption for floods since they are a sudden-onset weather event that tend to have instantaneous effects, but as droughts are slow-onset their negative effects on people’s livelihoods might not be seen for months or even years, depending on how long they last. This issue is addressed by the inclusion of the lagged effects variables in Tables 6 and 7.
These differing results could also be due to data limitations. One main limitation of this dataset is the fact that the flood and drought occurrences are measured at department level. Since there may be instances in which floods only affect one small area within a large department, the data cannot capture the fact that some people who are counted as being affected by this weather event by virtue of being in the same department might not be in the vicinity of the affected area at all. However, there is also the possibility that weather events in one part of the department can have consequences on other areas, through the mechanisms of higher food prices due to agricultural damage or limitations in human travel flows.
